---
title: "Data Visualization with ggplot2"
output:
        html_document:
            includes:
                in_header: header.html
            theme: paper
            toc: true
            number_sections: false
            fig_caption: false
            markdown: GFM
            code_download: false
---

<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 12px;
}
pre {
  font-size: 12px
}

</style>

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("06-")
```

```{r, echo = FALSE, eval = TRUE, purl = FALSE, message = FALSE}
## silently read in CSV file from data folder
library(tidyverse)
gapminder <- read.csv("../data/gapminder_data.csv")
gapminder <- as_tibble(gapminder) %>% mutate(total_gdp = pop * gdpPercap)
```

# Overview

In this lesson we will work with `ggplot2` and its "grammar of graphics" to create a variety of plots using the `gapminder` data we saw in the last lesson.

# Grammar of graphics

When people talk about `ggplot2` you may hear them use the phrase "grammar of graphics," which is a conception of statistical graphics developed by Leland Wilkinson. The main components of the grammar (in bold) are defined as follows in the book [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org).

> All plots are composed of the **data**, the information to visualize, and a **mapping**, the description of the data's variables are mapped to aesthetic attributes. There are five mapping components:
>
> - A **layer** is a collection of geometric elements and statistical transformations. Geometric elements, **geom**s for short, represent what you actually see in the plot: points, lines, polygons, etc. Statistical transformations, **stat**s for short, summarise the data: for example, binning and counting observations to create a histogram, or fitting a linear model.
> - **Scale**s map values in the data space to values in the aesthetic space. This includes the use of colour, shape or size. Scales also draw the legend and axes, which make it possible to read the original data values from the plot (an inverse mapping).
> - A **coord**, or coordinate system, describes how data coordinates are mapped to the plane of the graphic. It also provides axes and gridlines to help read the graph. We normally use the Cartesian coordinate system, but a number of others are available, including polar coordinates and map projections.
> - A **facet** specifies how to break up and display subsets of data as small multiples. This is also known as conditioning or latticing/trellising.
> - A **theme** controls the finer points of display, like the font size and background colour. While the defaults in ggplot2 have been chosen with care, you may need to consult other references to create an attractive plot.

Remembering all of this isn't necessary for generating plots quickly with `ggplot2`, so don't worry if these concepts aren't completely clear at first. As you work with `ggplot2` you will start to get a sense for how these things fit together and what function calls to use.

# Scatter plots

We first load the `ggplot2` library:

```{r load-ggplot2, message=FALSE, purl=FALSE}
library(ggplot2)
```

The plots you generate with `ggplot2` will usually conform to the template:

```{r, eval=FALSE}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +
        <GEOM_FUNCTION>() +
        <COORD_FUNCTION>() +
        <SCALE_FUNCTION>() +
        <THEME_FUNCTION>()
```

We may not always use coordinate, scale, and theme functions, but at minimum we **must** specify the data, mapping, and geometry functions. `ggplot2` functions need data in the 'long' format, i.e., a column for every dimension, and a row for every observation. Well-structured data will save you lots of time when making figures with `ggplot2`. The `gapminder` data that we have been using is already in the format best-suited for `ggplot2`, but keep in mind you might need to use the `dplyr` functions we learned in the last lesson to get to this form for other data you may encounter.

ggplots can be built iteratively, which is great for understanding what each addition does to the resulting plot.

1. Use the `ggplot()` function and bind the plot to the `gapminder` data frame with the `data` argument:

```{r ggplot_1, purl=FALSE}
ggplot(data = gapminder)
```

This plot is empty because we told it what data to use, but neither the mappings nor geometries.

2. Let's define a mapping (using the aesthetic (`aes`) function), by selecting the variables to be plotted and specifying how to present them in the graph. First we'll specify x/y axes, but later we will see how to use size, shape, color, etc.

```{r ggplot_2, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
```

This is a little better in that we at least have axes, but there is still no data on the plot because we didn't specify a geometry.

3. Now let's add a 'geom' – graphical representation of the data in the plot (points, lines, bars). `ggplot2` offers many different geoms; we will use some common ones today, including:
    * `geom_point()` for scatter plots, dot plots, etc.
    * `geom_line()` for trend lines, time series, etc.
    * `geom_barplot()` for, well, boxplots!
    * `geom_boxplot()` distributions.

To add a geom to the plot use the `+` operator (this is the `ggplot` version of the pipe from the previous lesson). Because we have two continuous variables, let's use `geom_point()` first:

```{r first-ggplot, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) + geom_point()
```

The `+` in the `ggplot2` package is particularly useful because it allows you to modify existing `ggplot` objects. Now we might want to change some things like the axis labels and add a title to the plot.

4. We can modify plot titles, axis labels, and their sizes, colors, etc. using theme functions. Let's give the axis labels better names, and give the plot a title using the `labs()` function.

```{r first-ggplot-labels, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy')
```

Now we have clearer labels and a good summary title. But what if we're not keen on that gray background? We can use a theme function to change the look and feel of the graph. The [Themes section](https://ggplot2-book.org/polishing.html) of the ggplot2 book provides a good overview of theming options.

5. Let's change the theme to something lighter:

```{r first-ggplot-theme, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
```

## Storing plots as objects

Base graphics in R typically don't allow plots to be saved and modified as objects, but you can with ggplots. This can be helpful for storing a base version of a plot to tinker with more easily without having to write some of the same code over and over again. For example:

```{r save-ggplot}
base_plot = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
```

And from here we can try some different things out, like changing the color of all the points:

```{r blue-ggplot}
base_plot + geom_point(color = 'blue')
```

Or we can make the points a little transparent:

```{r transparent-ggplot}
base_plot + geom_point(alpha = 0.5)
```

> ## Tip
>
> - Anything you put in the `ggplot()` function can be seen by any geom layers that you add (i.e., these are universal plot settings). This includes the x- and y-axis mapping you set up in `aes()`.
> - You can also specify mappings for a given geom independently of the mappings defined globally in the `ggplot()` function.
> - The `+` sign used to add new layers must be placed at the end of the line containing the *previous* layer.

## Exploring other aesthetics

### Color

To color each continent in the plot differently, you can specify that of the data to use as the `color` in the aesthetic (`aes()`) funciton.

```{r color-by-continent1, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
```

You can also change the colors that are used by specifying a named palette, or by manually defining the colors. [This page](https://r-graph-gallery.com/38-rcolorbrewers-palettes) shows some of the named palettes which are available, and `?scale_colour_brewer` also shows the names of the palettes.

```{r color-by-continent2, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    scale_color_brewer(palette = 'Set3') +
    theme_bw()
```

### Shape

You can also alter the shape of the points by specifying `shape` in the `aes()` function:

```{r shape-by-continent, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, shape = continent)) +
    geom_point() +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
```

### Size

You can alter the size of points by specifying `size` in the `aes()` function while also using the continents as a `color`. Let's save this plot as an object to use later in this lesson.

```{r size-by-pop, purl=FALSE}
gdp_life_plot <- ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
    geom_point(alpha = 0.5) +
    labs(title = 'GDP per Capita vs Life Expectancy (Size by Population)', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
# When saving a plot as an object, RStudio won't automatically display it
gdp_life_plot
```

### Faceting

`ggplot2` has a special technique called *faceting* that allows us to split one plot into multiple plots based on a factor in the data. Let's use it to split our plot into five panels, one for each continent.

```{r first-facet, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() +
    facet_grid(. ~ continent) +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
```

We can also experiment with stacking the facets vertically, rather than horizontally. The `facet_grid` geometry allows you to explicitly specify how you want your plots to be arranged via formula notation (`rows ~ columns`; a `.` can be used as a placeholder that indicates only one row or column).

```{r second-facet, purl=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() +
    facet_grid(continent ~ .) +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
```

# Line plots

We've explored a lot of options with the `geom_point()` geometry, but we should expand our horizons to other geometries. Let's first create a subset of the `gapminder` data limited only to Poland and plot the life expectancy over all the years for which we have data as a line plot.

> ## Exercise
>
> Use the `filter()` function to select only the Poland data from the `gapminder` data, and save it as as an object named "poland".

<details>
<summary>Solution</summary>
```{r subset-challenge, purl=FALSE}
poland = gapminder %>% filter(country == 'Poland')
poland
```
</details>
<br>

> ## Exercise
>
> For the `poland` subset of the `gapminder` data, plot the life expectancy as a function of the year. Make sure to label the axes and give the plot an appropriate title. Make sure to save your plot as an object called "poland_plot". Hint: Try the `geom_line()` geometry.

<details>
<summary>Solution</summary>
```{r lineplot-challenge, purl=FALSE}
poland_plot <- ggplot(data = poland, aes(x = year, y = lifeExp)) +
    geom_line() +
    labs(x = 'Year', y = 'Life Expectancy', title = 'Polish Life Expectancy Over Time')
poland_plot
```
</details>
<br>

This is excellent! Now let's consider plotting a similar line graph of life expectancy across time for **all countries** on the **same axes**. Moreover, let's color the lines by their continent. Perhaps we can even take as a guide our scatterplot from above:

```{r eval = FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() +
    labs(title = 'GDP per Capita vs Life Expectancy', x = 'GDP per Capita', y = 'Life Expectancy') +
    theme_bw()
```

We might just replace `gdpPercap` with `lifeExp` and `geom_point()` with `geom_line()` and hopefully get what we want. Let's give it a shot:

```{r first_line_plot, purl=FALSE}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color = continent)) +
    geom_line() +
    labs(title = 'Life Expectancy Over Time by Country', x = 'Year', y = 'Life Expectancy') +
    theme_bw()
```

This doesn't really look like what we expected. I expected a line for each country, but I can't really explain how ggplot decided to create the lines. There is a parameter of `aes()` called `group` which can help us. It will tell ggplot to first group the data by a column of the data, in this we probably want `country`. What happens if we try that? (Let's also save this plot as an object to use later.)

```{r second_line_plot, purl=FALSE}
life_over_time_plot <- ggplot(data = gapminder, aes(x = year, y = lifeExp, color = continent, group = country)) +
    geom_line() +
    labs(title = 'Life Expectancy Over Time by Country', x = 'Year', y = 'Life Expectancy') +
    theme_bw()
life_over_time_plot
```

Voila, we've got ourselves a spaghetti plot.

# Saving plots to file

Let's take a brief pause from experimenting with plots and learn how to save plots as image files. We should have three plots that we've saved as objects: `gdp_life_plot`, `poland_plot`, and `life_over_time_plot`. Let's save these plots as files using the `ggsave()` function.

The `ggsave()` function is an easy way to specify which plots to save, at what location, at what size and resolution, and in what format. Let's first have a look at the help file to figure out which parameters we should pay attention to.

```{r eval = FALSE}
?ggsave
```

The `filename` parameter can determine what format to write the file in based on the extension (e.g. `.tiff`, `.png`, `.pdf`, or `.jpeg`). Next the `height` and `width` can be specified in any number of `units` to get the exact size figure you prefer. Finally, the `dpi` parameter specifies the resolution in "dots per inch", and this parameter can ensure that the plots you output are sufficiently high-quality to submit as part of a publication to a journal. Typically 300dpi is sufficient. With this in mind, let's save the `gdp_life_plot`:

```{r save-plot, eval = FALSE}
ggsave(filename = 'gdp_life_plot.png', plot = gdp_life_plot, width = 6, height = 6, units = 'in', dpi = 300)
```

# Bar plots

What if we wanted to plot the global total GDP per year as a bar plot, but separated by continent? Let's first take a look at the help function for `geom_bar()`:

~~~
?geom_bar
~~~

Looking through the description:

- `geom_bar()` makes the height of the bar proportional to the number of cases in each group.
- If you want the heights of the bars to represent values in the data, use `geom_col()` instead.

Since we want the height of the bars to represent the values of the `total_gdp` column, we should be using `geom_col()`, a relative of `geom_bar()`. To make this graph, it's clear we want the `year` on the x-axis, the `total_gdp` on the y-axis, and if we want to separate the continents apart in the graph, we will use a new aesthetic parameter called `fill` and assign it `continent`.

```{r barplot1, purl=FALSE}
ggplot(data = gapminder, aes(x = year, y = total_gdp, fill = continent)) + geom_col()
```

This is a good start, and from previous examples we know how we might alter the x and y axis labels, but how can we edit the legend title? This is a common task when creating publication figures because the legend title will inherit the name of the column, which is usually some shorthand. We can use the same `labs()` function we used before, and specify a value for `fill`, to go with the `fill` parameter in `aes()` above. Let's save this plot as "total_gdp_plot".

```{r barplot2, purl=FALSE}
total_gdp_plot = ggplot(data = gapminder, aes(x = year, y = total_gdp, fill = continent)) +
    geom_col() +
    labs(title = 'Global GDP Over Time (by continent)', x = 'Year', y = 'Total GDP', fill = 'Continent')
total_gdp_plot
```

# Box plots

What if we wanted a plot to look at the distribution of populations of countries grouped by continents over the years? Let's start by building this plot iteratively.

```{r boxplot, purl = FALSE}
ggplot(data = gapminder, aes(x = continent, y = pop)) + geom_boxplot()
```

This is a good start, but because of the wide range of populations, we're having a problem seeing the distribution of values at the lower end. This is a perfect opportunity to change the scale of the y-axis to a logarithmic scale using the `scale_y_log10()` function.

```{r boxplot2, purl = FALSE}
ggplot(data = gapminder, aes(x = continent, y = pop)) + geom_boxplot() + scale_y_log10()
```

This is better, but we still have all the years lumped into each continent bar. We could split out the years as facets. We saw an example of that with `facet_grid()` above. Since we're trying to compare the distributions of populations across years at a glance, it would probably best to facet the years horizontally.

```{r boxplot3, purl = FALSE}
ggplot(data = gapminder, aes(x = continent, y = pop)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_grid(. ~ year)
```

This is great, we've come a long way, but maybe we'd actually like the x-axis to be by year and the facets to be by continent. That would feel like a better display of the data.

```{r boxplot4, purl = FALSE}
ggplot(data = gapminder, aes(x = year, y = pop)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_grid(. ~ continent)
```

That's an interesting warning. Because year is considered a numeric, we're not really getting what we'd like. ggplot is helpfully suggesting adding `group = year`.

```{r boxplot5, purl = FALSE}
ggplot(data = gapminder, aes(x = year, y = pop, group = year)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_grid(. ~ continent)
```

That feels like closer to what we intended. It also has the benefit of being beautiful in a way.

## Rotating axis labels

You will very likely discover in the course of making plots that the axis labels may collide with one another, and that rotating the labels would make for a better plot. Let's modify the plot we just made to rotate the years by 45 degrees. But where to start? Typically the answer for ggplot questions is to search the internet and look at example code. Let's do just that. Let's search for "how to rotate axis labels in ggplot".

...

It looks like a possible solution is to add

```{r, eval = FALSE}
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

Let's give it a shot:

```{r boxplot6, purl = FALSE}
ggplot(data = gapminder, aes(x = year, y = pop, group = year)) +
    geom_boxplot() +
    scale_y_log10() +
    facet_grid(. ~ continent) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

Perfect!

# Summary

In this lesson we've introduced the basic concepts underpinnning ggplot2:

- Data
- Mappings
- Layers
- Geoms
- Stats
- Scales
- Coords
- Facets
- Themes

We've introduced a number of specific geometries:

- `geom_point()`
- `geom_line()`
- `geom_bar()` and `geom_col()`
- `geom_boxplot()`

And we explored various customizations:

- Color palettes
- Plot, axis, and legend titles
- Axis label rotation

And we explored how to save our plots in different formats and resolutions to ensure publication quality graphics.

# Resources

* [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/introduction.html) is a definitive book of the Grammar of Graphics and `ggplot2` implementation.
* [ggplot2 official page](https://ggplot2.tidyverse.org/) is the official landing page describing ggplot2 as part of the tidyverse.
* [patchwork](https://patchwork.data-imaginist.com/), a package to easily add plots together!
